Fractional Types

### 2019-11-17 fracGC/p.tex

\paragraph*{Quantum and Reversible Computing.} The Toffoli gate is
known to be universal for classical reversible
circuits~\cite{Toffoli:1980}. Adding just one gate (the Hadamard gate)
produces a universal set of primitives for quantum
circuits~\cite{hadtoffuniv}. The ``only'' difference between the two
circuit models is that quantum circuits can process superpositions of
values (waves) in one step whereas classical circuits lack this form
of parallelism. Most importantly, structurally the two circuits models
are identical and one can derive properties valid for both families by
focusing on the simpler classical model. In fact, classical reversible
computations are often used as ``subroutines'' of quantum
computations.

Instead of using the Toffoli gate as a universal primitive for
reversible classical circuits, one can leverage the full force of type
theory and category theory by expressing reversible classical
computations as \emph{isomorphisms over finite
  types}~\cite{Fiore:2004,James:2012:IE:2103656.2103667} or
\emph{equivalences over
  groupoids}~\cite{DBLP:conf/esop/CaretteS16}. This perspective
exposes a rich mathematical structure in reversible computations
that we will exploit to solve the ``ancilla problem'' explained next.

\paragraph*{Temporary Storage using Ancilla Bits.} The universality of
the Toffoli gate for classical reversible computing and of the
universality of the combination of the Toffoli and Hadamard gates for
quantum computing should not distract from efficiency and safety
concerns. The theorems proving universality (i) assume that temporary
storage (often called \emph{ancilla bits}) may be
used~\cite{Toffoli:1980}, and (ii) that this temporary storage is
returned to its initial state before de-allocation. Indeed if no
temporary storage is allowed, the Toffoli gate is not
universal~\cite{DBLP:conf/innovations/AaronsonGS17} and as we
demonstrate in Sec.~\ref{sec:examples}, the more temporary storage is
allowed, the more efficient certain computations could become. The
condition requiring that the temporary storage is only de-allocated
when returned to its initial condition is a safety
condition. Violating this condition destroys the symmetry between
input and output making the circuits not reversible and, in the
quantum model, causes decoherence that may destroy the quantum
state. As reviewed in Sec.~\ref{sec:examples}, ancilla bits have a
number of critical applications and yet are poorly supported in
current reversible and quantum programming languages making them a
common source of bugs.

@@

Uses of ancillas:
\url{https://quantum.country/search}:

@@

\url{https://www.nap.edu/read/25196/chapter/5#73}
\begin{quote}
  For error correction, one needs to replicate the state of a qubit
  onto a number of qubits. While the no cloning theorem prevents one
  from copying the state of a qubit directly onto another, one can
  create a redundant entangled qubit state of many qubits. The key is
  that the qubits to be entangled must start out in a known
  state. Qubits with a known state (for purposes of this discussion,
  it will be the state $\ket{0}$), called ``ancilla qubits,'' may be
  added to a computation for this purpose. Since the state of ancilla
  qubits are known, it is possible to create a simple circuit that
  makes the output state of all these ancilla qubits match the
  protected qubit: run each ancilla through a controlled-NOT gate,
  where the control is driven by the qubit that needs to be
  replicated. Assume that there is a qubit with state $\Psi$ that we
  want to protect, where $\ket{\Psi}$ represents an arbitrary
  superposition state $\ket{\Psi} = a_0 \ket{0} + a_1 \ket{1}$. In the
  CNOT gate, the ancilla $\ket{0}$ state will remain a $\ket{0}$ state
  by the $\ket{0}$ component of $\ket{\Psi}$, but it will be converted
  to $\ket{1}$ by the $\ket{1}$ component of $\ket{\Psi}$. The result
  of this operation is the newly entangled two-qubit state
  $a_0 \ket{0} + a_1 \ket{1}$, creating a system in which the ancilla
  qubit is now perfectly entangled with the first qubit. Adding more
  ancillas increases the distance of the repetition code.  \end{quote}

@@

Toffoli4 using 2 Toffoli3: core of proof of universality; simple
enough. Reversible/Quantum Circuits and Ancilla Wires: Early use of
ancillas in Toffoli's paper to implement arbitrary reversible
functions using a fixed number of 3 input gates
\url{https://link.springer.com/content/pdf/10.1007%2F3-540-10003-2_104.pdf}

>>>>

\paragraph*{Quantum and Reversible Computing Based on Type
  Isomorphisms.} The Toffoli gate is known to be universal for
classical reversible circuits~\cite{Toffoli:1980}. Adding just one
gate (the Hadamard gate) produces a universal set of primitives for
quantum circuits~\cite{hadtoffuniv}. The ``only'' difference between
the two circuit models is that quantum circuits can process
superpositions of values (waves) in one step whereas classical
circuits lack this form of parallelism. Most importantly, structurally
the two circuits models are identical and one can derive properties
valid for both families by focusing on the simpler classical model. In
fact, classical reversible computations are often used as
``subroutines'' of quantum computations.

Instead of using the Toffoli gate as a universal primitive for
reversible classical circuits, one can leverage the full force of type
theory and category theory by expressing reversible classical
computations as \emph{isomorphisms over finite
  types}~\cite{Fiore:2004,James:2012:IE:2103656.2103667} or
\emph{equivalences over
  groupoids}~\cite{DBLP:conf/esop/CaretteS16}. This perspective is
similarly universal for reversible computation (and can be extended to
quantum computation~\cite{10.1007/978-3-319-89366-2_19}) but with the
advantage of exposing a rich mathematical structure in reversible
computations that we will exploit to solve the ``ancilla problem''
explained next.

\paragraph*{Temporary Storage using Ancilla Bits.} The universality of
the Toffoli gate for classical reversible computing and of the
universality of the combination of the Toffoli and Hadamard gates for
quantum computing should not distract from efficiency and safety
concerns. The theorems proving universality (i) assume that temporary
storage (often called \emph{ancilla bits}) may be
used~\cite{Toffoli:1980}, and (ii) that this temporary storage is
returned to its initial state before de-allocation. Indeed if no
temporary storage is allowed, the Toffoli gate is not
universal~\cite{DBLP:conf/innovations/AaronsonGS17} and as would be
expected the more temporary storage is allowed, the more efficient
certain computations could become. The condition requiring that the
temporary storage is only de-allocated when returned to its initial
condition is a safety condition. Violating this condition destroys the
symmetry between input and output making the circuits not reversible
and, in the quantum model, causes decoherence that may destroy the
quantum state. As reviewed in Sec.~\ref{sec:examples}, ancilla bits
have a number of critical applications and yet are poorly supported in
current reversible and quantum programming languages making them a
common source of bugs.

@@

Uses of ancillae:
\url{https://quantum.country/search}:

@@

\url{https://www.nap.edu/read/25196/chapter/5#73}
\begin{quote}
  For error correction, one needs to replicate the state of a qubit
  onto a number of qubits. While the no cloning theorem prevents one
  from copying the state of a qubit directly onto another, one can
  create a redundant entangled qubit state of many qubits. The key is
  that the qubits to be entangled must start out in a known
  state. Qubits with a known state (for purposes of this discussion,
  it will be the state $\ket{0}$), called ``ancilla qubits,'' may be
  added to a computation for this purpose. Since the state of ancilla
  qubits are known, it is possible to create a simple circuit that
  makes the output state of all these ancilla qubits match the
  protected qubit: run each ancilla through a controlled-NOT gate,
  where the control is driven by the qubit that needs to be
  replicated. Assume that there is a qubit with state $\Psi$ that we
  want to protect, where $\ket{\Psi}$ represents an arbitrary
  superposition state $\ket{\Psi} = a_0 \ket{0} + a_1 \ket{1}$. In the
  CNOT gate, the ancilla $\ket{0}$ state will remain a $\ket{0}$ state
  by the $\ket{0}$ component of $\ket{\Psi}$, but it will be converted
  to $\ket{1}$ by the $\ket{1}$ component of $\ket{\Psi}$. The result
  of this operation is the newly entangled two-qubit state
  $a_0 \ket{0} + a_1 \ket{1}$, creating a system in which the ancilla
  qubit is now perfectly entangled with the first qubit. Adding more
  ancillae increases the distance of the repetition code.  \end{quote}

@@

Toffoli4 using 2 Toffoli3: core of proof of universality; simple
enough. Reversible/Quantum Circuits and Ancilla Wires: Early use of
ancillae in Toffoli's paper to implement arbitrary reversible
functions using a fixed number of 3 input gates
\url{https://link.springer.com/content/pdf/10.1007%2F3-540-10003-2_104.pdf}
